---
title: "Data 607 Week 12 Assignment"
author: "Robert Lauto"
date: "11/25/2018"
output:
  html_document:
    toc: true
    toc_float: true

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(mongolite)
library(RMySQL)
```
# Migrating from RDB to MongoDB

The task of moving data can sometimes be daunting but with the help of R and packages like mongolite and RMySQL it is fairly easy.

## Creating collection in Mongo db 'flights' and creating connection
```{r Mongo conn, eval=T}
# creating collection in mongo flights db and connecting to it
airlines_col <- mongo(collection = 'airlines',
                      db = 'flights',
                      url = 'mongodb://localhost')

airports_col <- mongo(collection = 'airports',
                      db = 'flights',
                      url = 'mongodb://localhost')

flights_col <- mongo(collection = 'flights',
                      db = 'flights',
                      url = 'mongodb://localhost')

planes_col <- mongo(collection = 'planes',
                      db = 'flights',
                      url = 'mongodb://localhost')

weather_col <- mongo(collection = 'weather',
                      db = 'flights',
                      url = 'mongodb://localhost')
```
## Connect to MySQL
```{r connecting to MySQL, eval = T}
# connecting to MySQL flights db
flights_sql <- dbConnect(MySQL(), user='root', password='root12345', dbname='flights', host='localhost')

# getting dataframes from all tables flights db
airlines <- dbGetQuery(flights_sql, 'SELECT * FROM airlines')
airports <- dbGetQuery(flights_sql, 'SELECT * FROM airports')
flights <- dbGetQuery(flights_sql, 'SELECT * FROM flights')
planes <- dbGetQuery(flights_sql, 'SELECT * FROM planes')
weather <- dbGetQuery(flights_sql, 'SELECT * FROM weather')
```
Some quick tidying to remove new line character '\r'
```{r Tidying, eval = T}
# remove "\r" from end of line in airlines data frame
airlines$name <- sub('\r$','',airlines$name)
```
## Inserting the tables from MySQL to the Mongo collections
```{r insert, eval = T}
# insert tables into collections, dropping anything in the collection first
airlines_col$drop()
airlines_col$insert(airlines)

airports_col$drop()
airports_col$insert(airports)

flights_col$drop()
flights_col$insert(flights)

planes_col$drop()
planes_col$insert(planes)

weather_col$drop()
weather_col$insert(weather)
```
## Quick check to make sure all of the data is now store in the collections
```{r check, eval = T}
# checking that the collections now match the MySQL tables
airlines_col$count('{}') == nrow(airlines)

airports_col$count('{}') == nrow(airports)

flights_col$count('{}') == nrow(flights)

planes_col$count('{}') == nrow(planes)

weather_col$count('{}') == nrow(weather)
```
# Conclusion
- Storing data in relational databases, like MySQL, are useful for structured relational data, ensuring data integrity.
- However, NoSQL databases like MongoDB increase felxiblilty in database structure and is more easily scaled
- Once connections are established, the insertion of a table from MySQL to MongoDB collection can be done with one line like below:
```{r drop, eval=T, include=F}
airports_col$drop()
```
```{r one line, eval = T}
airports_col$insert(dbGetQuery(flights_sql, 'SELECT * FROM airports'))
```
